Mapping Cancer: the Potential of Cartograms and Alternative Map Displays

### 2020-05-22 Response/Response.Rmd

Thank you for your letter containing your feedback regarding our submision "Cartogram Mapping and its Application to Cancer Data Visualization", for inclusion in the Focused Issue on Spatial Patterns in Cancer Epidemiology in the Annals of Cancer Epidemiology.

This review paper presented an exploration of cancer mapping methods, focusing on variations on choropleths for online cancer atlases.

**Editor suggestion:**
**We would suggest that the aim of the manuscript could be phrased to highlight the current approaches and limitations of typical visualisation methods for cancer maps, along with outlining possible alternative methods and challenges in implementing these?**

We have restructured our original submission to provide a clear path from current approaches used, to the alternative approaches that should be explored. We highlight the limitations of using current display methods, to encourage the use of alternative display methods. We also present the challenges that may be faced when implementing the alternative designs.

### 2020-05-22 letter2/submission2.Rmd
aligned table 2, changed links to be 1a-1g

Several suggestions were made regarding Table 2: Common measures for reporting cancer information.
Additional columns have been included.

@@

As data becomes more sparse, either by increasing geographic resolution or stratifying by age/sex, obtaining reliable estimates becomes more difficult. Many atlases analyzing smaller areas use statistical modelling to produce estimates, although some continue to use simple calculations and suppress regions where estimates are unstable (e.g. (mention some)). XXX

>>>>

Several suggestions were made regarding Table 1: Common measures for reporting cancer information. this table was previously Table 2 and an additional column has been included to indicate the cancer atlas examples in Figure 1 that display each statistic.

@@

As data becomes more sparse, either by increasing geographic resolution or stratifying by age/sex, obtaining reliable estimates becomes more difficult. Many atlases analyzing smaller areas use statistical modelling to produce estimates, although some continue to use simple calculations and suppress regions where estimates are unstable.

### 2020-05-22 literature.Rmd
aligned table 2, changed links to be 1a-1g

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. A choropleth map subdivides these domains into administrative regions such as countries, states, or suburbs.  When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner - city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas. This study summarises current practices for cancer atlases and investigates the alternative map displays that could be used to accurately represent the distribution of cancer statistics across a population, as many cancer atlases lack appropriate displays for population statistics. It is recommended that alternative displays are included in cancer atlases. In addition, with the ease of today’s technology, user interaction with the displays is encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence, or filter by demographic variables.

@@

Epidemiologists and statisticians have developed a range of statistics to communicate the burden of cancer, and the choice of statistics used in maps has changed in recent decades. Table 1<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population, and age-adjusted to correct for the higher prevalence of cancers in older populations. As described by Howe [@HEDP], Englishman P. Stocks advanced the field of mortality statistics by introducing the standardized mortality ratios in the 1930s, which was an improvement on crude death rates.

@@

@@

Table describes the common measures used to present cancer incidence in atlases. The calculations used to produce these measures often are also applied to mortality statistics.The Excess Hazard Ratio and Standardized Incidence Ratio are additional measures that can be found in atlases, but are not used in the images displayed in Figure 1.

@@

| Fig. | Atlas | Data | Statistic | Source |
|------|------------|-----------|----------------------|------------------|
| a | The Environment and Health<br> Atlas of England and Wales<br> [@EnvEnglandWales2] | Sex: Women<br> Year: 2010<br> Cancer: Lung | Relative risk | [Office for National Statistics<br> (ONS) (England) and from the<br> Welsh Cancer Intelligence and<br> Surveillance Unit (WCISU)](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br> Incidence, Mortality and <br> <br> prevalence Worldwide<br> [@Globocan] | Sex: Both<br> Year: 2018<br> Cancer: All invasive | Age standardized<br> incidence rates<br> (per 100,000) | [World Health Organization's<br> International Agency for<br> Research on Cancer](https://www.iarc.fr/<br>  |
| c | Atlas of Cancer in Queensland<br>  [@QLDcancerAtlas] | Sex: Males<br> Year: 1998 - 2007<br> Cancer: Lung | Relative incidence<br> ratio | [Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br> [@ACA] | Sex: Males<br> Year: 2010 - 2014<br> Cancer: Lung | Average standardised<br> incidence rate | [Australian Cancer Atlas](https://atlas.cancer.org.au/) |
| e | United States Cancer<br> Statistics: An Interactive<br> Cancer Statistics Website<br>  | Sex: Both<br> Year: 2016<br> Cancer: All types | Incidence rate per<br> 100,000 | With data from state cancer<br> registries: <br> [Centers for Disease Control and Prevention](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br> Rates in Spain<br> [@cancerSpain] | Sex: Both<br> Year: 2004 - 2008<br> Cancer: All types | Relative risk | <br> [Map of cancer mortality rates<br> in Spain](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html |
| g | Atlas of Childhood<br> Cancer in Ontario<br> [@OntarioPediatric] | Sex: Both<br> Year: 1995 - 2004<br> Cancer: All types | Incidence rate per<br> 100,00 | <br> [The Pediatric Oncology Group<br> of Ontario Networked Information<br> System](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

<!-- 416 words -->
A review of modern cancer atlases [@roberts2019communication] identified 33 cancer atlases published between 2010 and 2016. Atlases published between 2016 and 2018 have also been considered. Each of these online atlases uses choropleth maps. All except one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships. Figure 1<!--\ref{fig:choropleth-grid-create}--> displays a subset of maps from these atlases, the selection varies in the geographies explored. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows Globocan 2018 [@Globocan] which explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide using estimates based on available country cancer registries. Many atlases allow users to view sex-specific distributions, however sex is displayed in side-by-side maps in the Map of Cancer Mortality Rates in Spain (Figure 1f)<!--\ref{fig:choropleth-grid-create}--> [@cancerSpain].

There is large variation in the resolution of the maps. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global information at a national level. The United States Cancer Statistics [@USInteractive] shows data aggregated at the state level (n = 51). The Environment and Health Atlas of England and Wales [@EnvEnglandWales2] (Figure 1<!--\ref{fig:choropleth-grid-create}-->a) shows the relative risk for women developing lung cancer  at a neighborhood (small-area) scale (n = 8850). The Australian Cancer Atlas (Figure 1<!--\ref{fig:choropleth-grid-create}-->d) shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 [@abs2016] (n = 2292). The Atlas of Cancer in Queensland (Figure 1<!--\ref{fig:choropleth-grid-create}-->c) shows a subset of the Australian Statistical Local Areas (SLAs) located in the state of Queensland within Australia [@QLDcancerAtlas] (n = 478).

Age-specific atlases are less common. Figure 1<!--\ref{fig:choropleth-grid-create}-->g displays Atlas of Childhood Cancer in Ontario, this communicates the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric].

@@

<!--
Choropleth maps can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the ‘potentially more important results in the more populous communities’ that are geographically smaller [@SE]. 
Inset maps like in Brisbane city in Figure Figure 1<!--\ref{fig:choropleth-grid-create}c of the state of Queensland are commonly used to reduce distorted interpretations, but it is not always an appropriate solution. As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.
-->

@@

In the contiguous cartogram map the state of California has become much larger because of the large population density. This draws attention to the densely populated North-East region and detracts from the less populated Mid West.

In the non-contiguous the state of California has remained closer to its original size than its surrounding states. The North-East states have remained closer to their geographical size, for Massachusetts and Connecticut. This draws attention to the densely populated North-East region and the sparse Mid West.

In Dorling cartograms the North-East states remain closer to their neighbors and are slightly displaced from their geographic location. It highlights the sparsity of the population in the Mid West by the distance between the circles at the geographic centroids.

In the Hexagon Tessellation it is simple to contrast the neighboring states however the North-East regions have been displaced from their geographic location. It highlights the sparsity of the population in the Mid West by the light yellow color, the Age-Adjusted rate in Kentucky is the darkest and its neighbors are similar.

>>>>

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. A choropleth map subdivides these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner - city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas. This study summarises current practices for cancer atlases and investigates the alternative map displays that could be used to accurately represent the distribution of cancer statistics across a population, as many cancer atlases lack appropriate displays for population statistics. It is recommended that alternative displays are included in cancer atlases for a perceptually accurate display of the distribution of the burden of cancer over the population. If it possible given the time and budget constraints, atlases should also present countries or regions containing hetergeneous spatial areas in an alternative display, in addition to the familiar choropleth map. With the ease of today’s technology, user interaction with the displays is also encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence, or filter by demographic variables.

@@

Epidemiologists and statisticians have developed a range of statistics to communicate the burden of cancer, and the choice of statistics used in maps has changed in recent decades. Table 1<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases. Mortality rates are commonly presented as relative rates of risk across the population, and age-adjusted to correct for the higher prevalence of cancers in older populations. As described by Howe [@HEDP], the Englishman P. Stocks advanced the field of mortality statistics by introducing the standardized mortality ratios in the 1930s, which was an improvement on crude death rates.

@@

Obtaining reliable estimates becomes more difficult as data becomes more sparse, either by increasing geographic resolution or stratifying by age/sex. Many atlases analyzing smaller areas such as the Australian Cancer Atlas [@ACA], use statistical modelling to produce estimates, although some continue to use simple calculations and suppress regions where estimates are unstable.

@@

@@

| Fig. | Atlas | Data | Statistic | Source |
|----|-----------|-------------|-----------|--------------|
| a | The Environment and Health<br> Atlas of England and Wales<br> [@EnvEnglandWales2] | Sex: Women<br> Year: 2010<br> Cancer: Lung | Relative risk | Office for National Statistics<br> (ONS) (England) and from the<br> Welsh Cancer Intelligence and<br> Surveillance Unit (WCISU) [1a](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br> Incidence, Mortality and <br> <br> prevalence Worldwide<br> [@Globocan] | Sex: Both<br> Year: 2018<br> Cancer: All invasive | Age standardized<br> incidence rates<br> (per 100,000) | World Health Organization's<br> International Agency for<br> Research on Cancer [1b](https://www.iarc.fr/<br>)  |
| c | Atlas of Cancer in Queensland<br>  [@QLDcancerAtlas] | Sex: Males<br> Year: 1998 - 2007<br> Cancer: Lung | Relative incidence<br> ratio | Queensland Cancer Registry [1c](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br> [@ACA] | Sex: Males<br> Year: 2010 - 2014<br> Cancer: Lung | Average standardised<br> incidence rate | Australian Cancer Atlas [1d](https://atlas.cancer.org.au/) |
| e | United States Cancer<br> Statistics: An Interactive<br> Cancer Statistics Website<br>  | Sex: Both<br> Year: 2016<br> Cancer: All types | Incidence rate per<br> 100,000 | With data from state cancer<br> registries: <br> Centers for Disease Control and Prevention [1e](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br> Rates in Spain<br> [@cancerSpain] | Sex: Both<br> Year: 2004 - 2008<br> Cancer: All types | Relative risk | <br> Map of cancer mortality rates<br> in Spain [1f](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) |
| g | Atlas of Childhood<br> Cancer in Ontario<br> [@OntarioPediatric] | Sex: Both<br> Year: 1995 - 2004<br> Cancer: All types | Incidence rate per<br> 100,00 | <br> The Pediatric Oncology Group<br> of Ontario Networked Information<br> System [1g](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

<!-- 416 words -->
A review of modern cancer atlases [@roberts2019communication] identified 33 cancer atlases published between 2010 and 2016. Atlases published between 2016 and 2018 have also been considered. Each of these online atlases uses choropleth maps. All except one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships. Figure 1<!--\ref{fig:choropleth-grid-create}--> displays a subset of maps from these atlases, the selection varies in the geographies explored. Figure 1b<!--\ref{fig:choropleth-grid-create}--> shows Globocan 2018 [@Globocan] which explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide using estimates based on available country cancer registries. Many atlases allow users to view sex-specific distributions, however sex is displayed in side-by-side maps in the Map of Cancer Mortality Rates in Spain (Figure 1f)<!--\ref{fig:choropleth-grid-create}--> [@cancerSpain].

There is large variation in the resolution of the maps. Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global information at a national level. The United States Cancer Statistics [@USInteractive] shows data aggregated at the state level (n = 51). The Environment and Health Atlas of England and Wales [@EnvEnglandWales2] (Figure 1a<!--\ref{fig:choropleth-grid-create}-->) shows the relative risk for women developing lung cancer  at a neighborhood (small-area) scale (n = 8850). The Australian Cancer Atlas (Figure 1d<!--\ref{fig:choropleth-grid-create}-->) shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 [@abs2016] (n = 2292). The Atlas of Cancer in Queensland (Figure 1c<!--\ref{fig:choropleth-grid-create}-->) shows a subset of the Australian Statistical Local Areas (SLAs) located in the state of Queensland within Australia [@QLDcancerAtlas] (n = 478).

Age-specific atlases are less common. Figure 1g<!--\ref{fig:choropleth-grid-create}--> displays Atlas of Childhood Cancer in Ontario, this communicates the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric].

@@

<!--
Choropleth maps can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the ‘potentially more important results in the more populous communities’ that are geographically smaller [@SE]. 
Inset maps like in Brisbane city in Figure 1c<!--\ref{fig:choropleth-grid-create} of the state of Queensland are commonly used to reduce distorted interpretations, but it is not always an appropriate solution. As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.
-->

@@

In the contiguous cartogram map (2a) the state of California has become much larger because of the large population density. This draws attention to the densely populated North-East region and detracts from the less populated Mid West.

In the non-contiguous (2b) the state of California has remained closer to its original size than its surrounding states. The North-East states have remained closer to their geographical size, for Massachusetts and Connecticut. This draws attention to the densely populated North-East region and the sparse Mid West.

In the Dorling cartogram (2c) the North-East states remain closer to their neighbors and are slightly displaced from their geographic location. It highlights the sparsity of the population in the Mid West by the distance between the circles at the geographic centroids.

In the Hexagon Tessellation (2d) it is simple to contrast the neighboring states however the North-East regions have been displaced from their geographic location. It highlights the sparsity of the population in the Mid West by the light yellow color, the Age-Adjusted rate in Kentucky is the darkest and its neighbors are similar.
